---
title: "NPVegData Analysis 2025"
output:
  html_document: default
  pdf_document: default
date: "2025-10-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`````{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(readr)
library(readxl)
library(gt)
library(purrr)
`````

# Housekeeping

## Load Data

```{r}
#This year's data
alldata_now <- read_csv("2026-ALL.csv")

#Past data
alldata_25 <- read_csv("2025-ALL.csv")
alldata_23 <- read_csv("2023-ALL.csv")
alldata_22 <- read_csv("2022-ALL.csv")
alldata_21 <- read_csv("2021-ALL.csv")

```

## Clean data

```{r data clean}
#read in Name change csv which is used to standardize transect ID names
name_change <- read_csv("NameChange.csv") %>%
  clean_names()   # -> startid, endid



#Standardize cover_type to title case and remove extra spaces. Change transect IDs
alldata_now <- alldata_now %>%
  clean_names() %>%
  mutate(cover_type = str_to_title(str_trim(cover_type))) %>%
  left_join(name_change, by = c("transect" = "startid")) %>%
  mutate(transect = coalesce(endid, transect)) %>%
  select(-endid)

alldata_25 <- alldata_25 %>%
  clean_names() %>%
  mutate(cover_type = str_to_title(str_trim(cover_type)))%>%
  left_join(name_change, by = c("transect" = "startid")) %>%
  mutate(transect = coalesce(endid, transect)) %>%
  select(-endid)

alldata_23 <- alldata_23 %>%
  clean_names() %>%
  mutate(cover_type = str_to_title(str_trim(cover_type)))%>%
  left_join(name_change, by = c("transect" = "startid")) %>%
  mutate(transect = coalesce(endid, transect)) %>%
  select(-endid)

alldata_22 <- alldata_22 %>%
  clean_names() %>%
  mutate(cover_type = str_to_title(str_trim(cover_type)))%>%
  left_join(name_change, by = c("transect" = "startid")) %>%
  mutate(transect = coalesce(endid, transect)) %>%
  select(-endid)

alldata_21 <- alldata_21 %>%
  clean_names() %>%
  mutate(cover_type = str_to_title(str_trim(cover_type)))%>%
  left_join(name_change, by = c("transect" = "startid")) %>%
  mutate(transect = coalesce(endid, transect)) %>%
  select(-endid)

```

# Current Year - Core Metrics

## Percent Cover

```{r}
##Gross Cover

# Calculate cover for each category
native_cover <- alldata_now %>%
  filter(cover_type == "Native") %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

exotic_cover <- alldata_now %>%
  filter(cover_type == "Exotic") %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

# Calculate cover for specific subcategories based on species column
thatch_cover <- alldata_now %>%
  filter(species == "Thatch") %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

mowed_thatch_cover <- alldata_now %>%
  filter(species == "Mowed_thatch") %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

bare_ground_cover <- alldata_now %>%
  filter(species == "Bare_ground") %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

# Calculate remaining "Other" (cover_type = "Other" but not the three specific species)
remaining_other_cover <- alldata_now %>%
  filter(cover_type == "Other" & !species %in% c("Thatch", "Mowed_thatch", "Bare_ground")) %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

# Total cover
total_cover <- alldata_now %>%
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

# Calculate percentages
percent_native <- (native_cover / total_cover) * 100
percent_exotic <- (exotic_cover / total_cover) * 100
percent_thatch <- (thatch_cover / total_cover) * 100
percent_mowed_thatch <- (mowed_thatch_cover / total_cover) * 100
percent_bare_ground <- (bare_ground_cover / total_cover) * 100
percent_remaining_other <- (remaining_other_cover / total_cover) * 100


## Relative Cover

# Total vegetative cover (exclude 'Other')
total_veg_cover <- alldata_now %>%
  filter(cover_type != "Other") %>%   # only actual vegetation
  summarize(total = sum(cover, na.rm = TRUE)) %>%
  pull(total)

# Relative percentages
percent_relative_native <- (native_cover / total_veg_cover) * 100
percent_relative_exotic <- (exotic_cover / total_veg_cover) * 100


## Results

# Create a tibble with all categories
total_cover_pct <- tibble(
  Cover_Type = c("Native", "Exotic", "Other", "Other", "Other", "Other"),
  Subcategory = c("Native", "Exotic", "Thatch", "Mowed Thatch", "Bare Ground", "Other"),
  Percent_Cover = c(percent_native, percent_exotic, percent_thatch, 
                    percent_mowed_thatch, percent_bare_ground, percent_remaining_other)
)

# Plot it
ggplot(total_cover_pct, aes(x = Cover_Type, y = Percent_Cover, fill = Subcategory)) +
  geom_col() +
  labs(
    title = "Percent Vegetation Cover by Type",
    y = "Percent Cover (%)",
    x = "",
    fill = "Cover Category"
  ) +
  theme_minimal() +
    geom_text(aes(label = round(Percent_Cover, 1)),
            position = position_stack(vjust = 0.5),
            size = 3.2,
            color = "white") +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold")
  ) +
  scale_fill_manual(
    values = c("Native" = "forestgreen", 
               "Exotic" = "firebrick", 
               "Thatch" = "tan",
               "Mowed Thatch" = "tan4",
               "Bare Ground" = "saddlebrown",
               "Other" = "gray"),
    breaks = c("Native", "Exotic", "Thatch", "Mowed Thatch", "Bare Ground", "Other"),
    labels = c("Native", "Exotic", "  Thatch", "  Mowed Thatch", "  Bare Ground", "  Other")
  ) +
  guides(fill = guide_legend(
    title = "Cover Category",
    override.aes = list(size = 5)
  ))

#Create tibble for relative percent cover

relative_cover_pct <- tibble(
Cover_Type = c("Native", "Exotic"),
Percent_Cover = c(percent_relative_native, percent_relative_exotic)
)

#Plot relative percent cover

ggplot(relative_cover_pct, aes(x = Cover_Type, y = Percent_Cover, fill = Cover_Type)) +
  geom_col() +
  geom_text(
    aes(
      y = Percent_Cover / 2,               # place label halfway up the bar
      label = round(Percent_Cover, 1)
    ),
    color = "white",
    size = 3.2
  ) +
  labs(
    title = "Relative Vegetation Cover (% of Total Vegetative Cover)",
    y = "Percent Cover (%)",
    x = ""
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Native" = "forestgreen", "Exotic" = "firebrick"))


```

## Percent Cover By Transect

```{r}
# Get all unique transects

transects <- unique(alldata_now$transect)

# Initialize empty tibble for results

all_transect_summary <- tibble()

# Loop through each transect

for (VP in transects) {

# Filter data for this transect

VP_data <- alldata_now %>%
filter(transect == VP)

# Total cover

total_cover_VP <- sum(VP_data$cover, na.rm = TRUE)

# Native cover

native_cover_VP <- VP_data %>%
filter(cover_type == "Native") %>%
summarize(total = sum(cover, na.rm = TRUE)) %>%
pull(total)

# Exotic cover

exotic_cover_VP <- VP_data %>%
filter(cover_type == "Exotic") %>%
summarize(total = sum(cover, na.rm = TRUE)) %>%
pull(total)

# Other cover

other_cover_VP <- total_cover_VP - (native_cover_VP + exotic_cover_VP)

# Percent covers

percent_native_VP <- (native_cover_VP / total_cover_VP) * 100
percent_exotic_VP <- (exotic_cover_VP / total_cover_VP) * 100
percent_other_VP <- (other_cover_VP / total_cover_VP) * 100

# Relative cover (exclude Other)

total_veg_cover_VP <- native_cover_VP + exotic_cover_VP
percent_relative_native_VP <- (native_cover_VP / total_veg_cover_VP) * 100
percent_relative_exotic_VP <- (exotic_cover_VP / total_veg_cover_VP) * 100

# Store results

VP_summary <- tibble(
Transect = VP,
Percent_Native = percent_native_VP,
Percent_Exotic = percent_exotic_VP,
Percent_Other = percent_other_VP,
Relative_Native = percent_relative_native_VP,
Relative_Exotic = percent_relative_exotic_VP
)

# Append

all_transect_summary <- bind_rows(all_transect_summary, VP_summary)
}

##Graph it 

# Reshape for plotting

all_transect_long <- all_transect_summary %>%
pivot_longer(
cols = c(Percent_Native, Percent_Exotic, Percent_Other),
names_to = "Cover_Type",
values_to = "Percent_Cover"
) %>%
mutate(Cover_Type = case_when(
Cover_Type == "Percent_Native" ~ "Native",
Cover_Type == "Percent_Exotic" ~ "Exotic",
Cover_Type == "Percent_Other" ~ "Other"
))

# Set stacking order

all_transect_long$Cover_Type <- factor(all_transect_long$Cover_Type, levels = c("Other", "Exotic", "Native"))

# Plot stacked bars

ggplot(all_transect_long, aes(x = Transect, y = Percent_Cover, fill = Cover_Type)) +
geom_col() +
labs(
title = "Stacked Percent Cover by Transect",
y = "Percent Cover (%)",
x = "Transect"
) +
scale_fill_manual(values = c("Native" = "forestgreen", "Exotic" = "firebrick", "Other" = "gray")) +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  geom_text(aes(label = round(Percent_Cover, 1)),  # round to 1 decimal
            position = position_stack(vjust = 0.5), # place labels in the middle of each bar
            color = "white", size = 4)


# Relative cover (Native and Exotic only)

relative_cover_long <- all_transect_summary %>%
select(Transect, Relative_Native, Relative_Exotic) %>%
pivot_longer(
cols = c(Relative_Native, Relative_Exotic),
names_to = "Cover_Type",
values_to = "Percent_Cover"
) %>%
mutate(Cover_Type = case_when(
Cover_Type == "Relative_Native" ~ "Native",
Cover_Type == "Relative_Exotic" ~ "Exotic"
))

# Set order

relative_cover_long$Cover_Type <- factor(relative_cover_long$Cover_Type, levels = c("Exotic", "Native"))

# Plot

ggplot(relative_cover_long, aes(x = Transect, y = Percent_Cover, fill = Cover_Type)) +
  geom_col() +
  geom_text(aes(label = round(Percent_Cover, 1)),  # round to 1 decimal
            position = position_stack(vjust = 0.5), # place labels in the middle of each bar
            color = "white", size = 4) +
  labs(
    title = "Relative Vegetation Cover (% of Total Vegetative Cover)",
    y = "Percent Cover (%)",
    x = "Transect"
  ) +
  scale_fill_manual(values = c("Native" = "forestgreen", "Exotic" = "firebrick")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

## Problem weeds

```{r}
# Top 10 most abundant exotic species by total cover and percent of exotic cover

total_exotic_cover <- sum(alldata_now$cover[alldata_now$cover_type == "Exotic"], na.rm = TRUE)

top_exotic_species <- alldata_now %>%
filter(cover_type == "Exotic") %>%
group_by(species) %>%
summarize(Total_Cover = sum(cover, na.rm = TRUE)) %>%
mutate(Percent_of_Exotic = (Total_Cover / total_exotic_cover) * 100) %>%
arrange(desc(Total_Cover)) %>%
slice_head(n = 10)

##Listed
top_exotic_species

##Graphed

ggplot(top_exotic_species, 
       aes(x = reorder(species, Percent_of_Exotic),
           y = Percent_of_Exotic)) +
  geom_col(fill = "firebrick3") +
  coord_flip() +
  labs(
    title = "Top 10 Exotic Species by Percent of Exotic Cover",
    x = "Species",
    y = "Percent of Exotic Cover (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 9),
    plot.title = element_text(face = "bold")
  )




```

# Past Data - Core metrics

## Percent Cover

```{r}

# ---- MASTER DATA LIST[ADD NEW YEAR HERE] ----
all_data_list <- list(
  `2021` = alldata_21,
  `2022` = alldata_22,
  `2023` = alldata_23,
  
  `2025` = alldata_now    # replace with alldata_26 next year etc.
)

# ------ AUTOMATED CALCULATIONS AND GRAPHING BELOW  ---------

calc_cover_stats <- function(df) {
  
  total_cover  <- sum(df$cover, na.rm = TRUE)
  native_cover <- sum(df$cover[df$cover_type == "Native"], na.rm = TRUE)
  exotic_cover <- sum(df$cover[df$cover_type == "Exotic"], na.rm = TRUE)
  other_cover  <- total_cover - (native_cover + exotic_cover)

  # Percentages of total
  percent_native <- 100 * native_cover / total_cover
  percent_exotic <- 100 * exotic_cover / total_cover
  percent_other  <- 100 * other_cover  / total_cover

  # Relative percentages (no "other")
  veg_total <- native_cover + exotic_cover
  percent_relative_native <- 100 * native_cover / veg_total
  percent_relative_exotic <- 100 * exotic_cover / veg_total
  
  tibble(
    Percent_Native = percent_native,
    Percent_Exotic = percent_exotic,
    Percent_Other  = percent_other,
    Relative_Native = percent_relative_native,
    Relative_Exotic = percent_relative_exotic
  )
}


# Automatically run for all years 
all_years_summary <- imap_dfr(
  all_data_list,
  ~ calc_cover_stats(.x) %>% mutate(year = as.integer(.y))
)

# Create singular combined data set 
alldata_all <- bind_rows(
  imap(all_data_list, ~ mutate(.x, year = as.integer(.y)))
)

#Results

# Results Table
all_years_summary

#Results Graphs

# Combine all your datasets (2021–2025)
alldata_all <- bind_rows(alldata_21, alldata_22, alldata_23, alldata_now) %>%
  filter(!is.na(year))


# Summarize total cover per cover type per year
cover_summary <- alldata_all %>%
  group_by(year, cover_type) %>%
  summarize(total_cover = sum(cover, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  mutate(percent_cover = (total_cover / sum(total_cover)) * 100)

# Order cover types for consistent stacking
cover_summary$cover_type <- factor(
  cover_summary$cover_type,
  levels = c("Native", "Exotic", "Other")
)

# Plot 1: Total Percent Cover-----
ggplot(cover_summary, aes(x = factor(year), y = percent_cover, fill = cover_type)) +
  geom_bar(stat = "identity") +
    geom_text(aes(label = round(percent_cover, 1)),
            position = position_stack(vjust = 0.5),
            size = 3.2,
            color = "white") +
  labs(
    title = "Total Percent Cover by Year (2021–2025)",
    x = "Year",
    y = "Percent Cover",
    fill = "Cover Type"
  ) +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("forestgreen", "firebrick", "gray70"))

# Plot 2: Relative Percent Cover (Native vs Exotic only) ----
relative_summary <- cover_summary %>%
  filter(cover_type %in% c("Native", "Exotic")) %>%
  group_by(year) %>%
  mutate(percent_relative = (total_cover / sum(total_cover)) * 100)

ggplot(relative_summary, aes(x = factor(year), y = percent_relative, fill = cover_type)) +
  geom_bar(stat = "identity") +
    geom_text(aes(label = round(percent_relative, 1)),
            position = position_stack(vjust = 0.5),
            size = 3.2,
            color = "white") +
  labs(
    title = "Percent of Vegetation Cover by Year (2021–2025, Native vs Exotic)",
    x = "Year",
    y = "Percent Cover (Relative to Vegetation Only)",
    fill = "Cover Type"
  ) +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("forestgreen", "firebrick"))



```

## Species Richness

```{r}
# 2025 species counts
species_counts_25 <- alldata_now %>%
  filter(!species %in% c("Thatch", "Mowed_thatch", "Bare_ground")) %>%
  group_by(cover_type) %>%
  summarize(n_species = n_distinct(species)) %>%
  mutate(year = 2025)

# 2023 species counts
species_counts_23 <- alldata_23 %>%
  filter(!species %in% c("Thatch", "Mowed_thatch", "Bare_ground")) %>%
  group_by(cover_type) %>%
  summarize(n_species = n_distinct(species)) %>%
  mutate(year = 2023)

# 2022 species counts
species_counts_22 <- alldata_22 %>%
  filter(!species %in% c("Thatch", "Mowed_thatch", "Bare_ground")) %>%
  group_by (cover_type) %>%
  summarize(n_species = n_distinct(species)) %>%
  mutate(year = 2022)

# 2021 species counts
species_counts_21 <- alldata_21 %>%
  filter(!species %in% c("Thatch", "Mowed_thatch", "Bare_ground")) %>%
  group_by(cover_type) %>%
  summarize(n_species = n_distinct(species)) %>%
  mutate(year = 2021)

# Combine all years into one tibble
all_species_counts <- bind_rows(
  species_counts_21, species_counts_22, 
  species_counts_23, species_counts_25
)

# ---- Plot: Stacked Species Counts (Native + Exotic Only) ----

species_counts_no_other <- all_species_counts %>%
  filter(cover_type %in% c("Native", "Exotic"))

ggplot(species_counts_no_other, 
       aes(x = factor(year), y = n_species, fill = cover_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n_species),
            position = position_stack(vjust = 0.5),
            size = 3.4,
            color = "white") +
  labs(
    title = "Number of Species by Cover Type (2021–2025, excluding Other)",
    x = "Year",
    y = "Species Count",
    fill = "Cover Type"
  ) +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("forestgreen", "firebrick"))

```

## Problem Species over time

```{r}

calc_top5_exotics <- function(df) {
  df_ex <- df %>% filter(cover_type == "Exotic")
  total_ex <- sum(df_ex$cover, na.rm = TRUE)
  
  df_ex %>%
    group_by(species) %>%
    summarize(total_cover = sum(cover, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(total_cover)) %>%
    slice_head(n = 5) %>%
    mutate(percent_of_exotic = 100 * total_cover / total_ex)
}

top5_trends <- imap_dfr(all_data_list, ~ {
  calc_top5_exotics(.x) %>% mutate(year = as.integer(.y))
})

# Plot----------------------
ggplot(top5_trends, aes(x = year, y = percent_of_exotic, color = species, group = species)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Top 5 Exotic Species Each Year: Percent of Total Exotic Cover",
    x = "Year",
    y = "Percent of Exotic Cover (%)",
    color = "Species"
  ) +
  theme_minimal(base_size = 13) +
  scale_x_continuous(breaks = c(2021, 2022, 2023, 2025)) +
  scale_color_brewer(palette = "Set1") +
  ylim(0, max(top5_trends$percent_of_exotic) * 1.1)
```

# Playground and Deep Dive

## Lost and gained native species over time

```{r}
# Get native species from 2022
native_2022 <- alldata_22 %>%
  filter(cover_type == "Native") %>%
  distinct(species) %>%
  pull(species)

# Get native species from 2025
native_2025 <- alldata_now %>%
  filter(cover_type == "Native") %>%
  distinct(species) %>%
  pull(species)

# Find species in 2022 but not in 2025
species_lost <- setdiff(native_2022, native_2025)

species_lost

# Species in 2025 but not in 2022
species_gained <- setdiff(native_2025, native_2022)

species_gained
```

## Problem Transects Over Time

### Up1, Wet 4, Bio: Native vs Exotic Composition

```{r}

# Filter for the two transects of interest
focus_transects <- alldata_all %>%
  filter(transect %in% c("Up1", "Bio", "Wet4"))

# Summarize percent cover by year × transect × cover type
upbio_summary <- focus_transects %>%
  group_by(year, transect, cover_type) %>%
  summarize(total_cover = sum(cover, na.rm = TRUE), .groups = "drop") %>%
  group_by(year, transect) %>%
  mutate(percent_cover = (total_cover / sum(total_cover)) * 100)

# Only keep Native + Exotic for relative cover
upbio_relative <- upbio_summary %>%
  filter(cover_type %in% c("Native", "Exotic")) %>%
  group_by(year, transect) %>%
  mutate(percent_relative = (total_cover / sum(total_cover)) * 100)

### ---- Plot 1: Percent Cover Over Time (Up1, Wet4, Bio) ----
ggplot(upbio_summary, aes(x = year, y = percent_cover,
                          fill = cover_type)) +
  geom_col(position = "stack") +
  facet_wrap(~ transect) +
  labs(
    title = "Total Percent Cover Over Time (Up1 & Bio)",
    y = "Percent Cover",
    x = "Year",
    fill = "Cover Type"
  ) +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c(Native = "forestgreen",
                               Exotic = "firebrick",
                               Other = "gray70"))

### ---- Plot 2: Relative Cover (Native vs Exotic Only) ----
ggplot(upbio_relative,
       aes(x = year, y = percent_relative, fill = cover_type)) +
  geom_col(position = "stack") +
  facet_wrap(~ transect) +
  labs(
    title = "Relative Vegetation Cover (Native vs Exotic Only)",
    y = "Percent of Vegetative Cover",
    x = "Year",
    fill = "Cover Type",
  ) +
  theme_minimal(base_size = 13) +
  theme(
  axis.text.x = element_text(size = 8)   # or whatever size you want
) +
  scale_fill_manual(values = c(Native = "forestgreen",
                               Exotic = "firebrick"))
```

### Problem Transects Cont. : Native Cover Loss over time

```{r}

# Wet4 native cover: 2022 vs 2025
wet4_native_22 <- alldata_22 %>%
  filter(transect == "Wet4", cover_type == "Native") %>%
  group_by(species) %>%
  summarize(total_cover_2022 = sum(cover, na.rm = TRUE), .groups = "drop")

wet4_native_25 <- alldata_now %>%
  filter(transect == "Wet4", cover_type == "Native") %>%
  group_by(species) %>%
  summarize(total_cover_2025 = sum(cover, na.rm = TRUE), .groups = "drop")

wet4_native_change <- full_join(wet4_native_22, wet4_native_25, by = "species") %>%
  replace_na(list(total_cover_2022 = 0, total_cover_2025 = 0)) %>%
  mutate(
    cover_change = total_cover_2025 - total_cover_2022,
    percent_change = ifelse(total_cover_2022 == 0, NA,
                            (cover_change / total_cover_2022) * 100)
  ) %>%
  arrange(cover_change) %>%  # largest negative first
  slice_head(n = 5)


# Up1 native cover: 2022 vs 2023
up1_native_22 <- alldata_22 %>%
  filter(transect == "Up1", cover_type == "Native") %>%
  group_by(species) %>%
  summarize(total_cover_2022 = sum(cover, na.rm = TRUE), .groups = "drop")

up1_native_23 <- alldata_23 %>%
  filter(transect == "Up1", cover_type == "Native") %>%
  group_by(species) %>%
  summarize(total_cover_2023 = sum(cover, na.rm = TRUE), .groups = "drop")

# Combine and calculate change
up1_native_change_22_23 <- full_join(up1_native_22, up1_native_23, by = "species") %>%
  replace_na(list(total_cover_2022 = 0, total_cover_2023 = 0)) %>%
  mutate(
    cover_change = total_cover_2023 - total_cover_2022,   # negative = drop
    percent_change = ifelse(total_cover_2022 == 0, NA,
                            (cover_change / total_cover_2022) * 100)
  ) %>%
  arrange(cover_change) %>%  # most negative = biggest drop
  slice_head(n = 5)          # top 5 losses

# Create single table with all loss for 2023 and 2025 --------------------

# Step 1: Sum cover per species per year (Up1 only)
# -------------------------
up1_cover <- bind_rows(
  alldata_22 %>%
    filter(transect == "Up1", cover_type == "Native") %>%
    group_by(species) %>%
    summarize(cover = sum(cover, na.rm = TRUE), .groups = "drop") %>%
    mutate(year = 2022),

  alldata_23 %>%
    filter(transect == "Up1", cover_type == "Native") %>%
    group_by(species) %>%
    summarize(cover = sum(cover, na.rm = TRUE), .groups = "drop") %>%
    mutate(year = 2023),

  alldata_now %>%
    filter(transect == "Up1", cover_type == "Native") %>%
    group_by(species) %>%
    summarize(cover = sum(cover, na.rm = TRUE), .groups = "drop") %>%
    mutate(year = 2025)
)

# -------------------------
# Step 2: Pivot wider so each year is a column
# -------------------------
up1_cover_wide <- up1_cover %>%
  pivot_wider(
    names_from = year,
    values_from = cover,
    names_prefix = "cover_",
    values_fill = 0
  ) %>%
  # Step 3: Calculate losses relative to 2022
  mutate(
    loss_2023 = cover_2023 - cover_2022,
    loss_2025 = cover_2025 - cover_2022
  ) %>%
  # Step 4: Keep only species with loss >= 50
  filter(loss_2023 <= -50 | loss_2025 <= -50) %>%
  # Step 5: Sort by largest 2025 loss
  arrange(loss_2025)

# -------------------------
# Step 6: Display table with highlighted loss columns
# -------------------------
up1_cover_wide %>%
  gt() %>%
  cols_label(
    species = "Species",
    cover_2022 = "2022 Cover",
    cover_2023 = "2023 Cover",
    cover_2025 = "2025 Cover",
    loss_2023 = "2023 Loss",
    loss_2025 = "2025 Loss"
  ) %>%
  fmt_number(
    columns = vars(cover_2022, cover_2023, cover_2025, loss_2023, loss_2025),
    decimals = 1
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "pink"),
    locations = gt::cells_body(columns = vars(loss_2023, loss_2025))
  )

```
